Predicting Federal Emergency Management Agency Funding for U.S. Disasters

### 2021-11-13 code/analysis_code/analysis-eda.Rmd
Manuscript updates

#create a table where columns represent requested and obligated funds and rows are 5 most disaster-prone states
decs_state_5 <- EDAdata %>%
                    dplyr::add_count(state) %>%
                    dplyr::filter(dense_rank(-n) < 6) %>%
                    gtsummary::select(state, ReqAmt, OblAmt) %>%
                    gtsummary::tbl_summary(
                      by = state,
                      label = list(state ~ "State or Territory",
                                   ReqAmt ~ "Requested Funds",
                                   OblAmt ~ "Obligated Funds"),
                      statistic = list(all_continuous() ~ "{mean} ({sd})")) %>%
                    gtsummary::modify_header(label ~ "**Variable**") %>% 
                    gtsummary::modify_spanning_header(c("stat_1", "stat_2", "stat_3", "stat_4", "stat_5") ~ "**State**") %>%
                    gtsummary::modify_caption("**Table 1. FEMA Funding for 5 Most Disaster-Prone States 2012 - 2021**") %>%
                    gtsummary::bold_labels()
decs_state_5

>>>>

#create a table where columns represent requested and obligated funds and rows are 5 most disaster-prone states
decs_state_5 <- EDAdata %>%
                    dplyr::add_count(state) %>%
                    dplyr::filter(dense_rank(-n) < 6) %>%
                    gtsummary::select(state, ReqAmt, OblAmt) %>%
                    gtsummary::tbl_summary(
                      by = state,
                      label = list(state ~ "State or Territory",
                                   ReqAmt ~ "Requested Funds",
                                   OblAmt ~ "Obligated Funds"),
                      statistic = list(all_continuous() ~ "{mean} ({sd})")) %>%
                    gtsummary::modify_header(label ~ "**Variable**") %>% 
                    gtsummary::modify_spanning_header(c("stat_1", "stat_2", "stat_3", "stat_4", "stat_5") ~ "**State**") %>%
                    gtsummary::modify_caption("**Table 4. FEMA Funding for 5 Most Disaster-Prone States 2012 - 2021**") %>%
                    gtsummary::bold_labels()
decs_state_5

### 2021-11-13 code/analysis_code/analysis-linearmodels.Rmd
Manuscript updates

Create box and whisker plot for the fit output.
```{r req SLR bp}
req_SLR_bp <- broom.mixed::tidy(req_SLR_fit) %>%
                dotwhisker::dwplot(dot_args = list(size = 2, color = "blue"),
                                   whisker_args = list(color = "blue"),
                                   vline = geom_vline(xintercept = 0, colour = "grey50", linetype = 2))
req_SLR_bp
```

@@

#box and whisker plot for lm_fit2 significant predictors
req_MVR_fit_bp2 <- req_MVR_fit_sig %>%
                        dotwhisker::dwplot(dot_args = list(size = 2, color = "blue"),
                                           whisker_args = list(color = "blue"),
                                           vline = geom_vline(xintercept = 0, colour = "grey50", linetype = 2))
req_MVR_fit_bp2
```
This model identifies total number of federal agencies, the state/territory being Puerto Rico, and emergency declarations as significant predictors of requested FEMA funding. It makes sense that an emergency declaration is a protective factor, as it is a smaller event than a major disaster declaration.

@@

Create box and whisker plot for the fit output.
```{r obl SLR bp}
obl_SLR_bp <- broom.mixed::tidy(obl_SLR_fit) %>%
                dotwhisker::dwplot(dot_args = list(size = 2, color = "blue"),
                                   whisker_args = list(color = "blue"),
                                   vline = geom_vline(xintercept = 0, colour = "grey50", linetype = 2))
obl_SLR_bp
```

@@

#box and whisker plot for lm_fit2 significant predictors
obl_MVR_fit_bp2 <- obl_MVR_fit_sig %>%
                        dotwhisker::dwplot(dot_args = list(size = 2, color = "blue"),
                                           whisker_args = list(color = "blue"),
                                           vline = geom_vline(xintercept = 0, colour = "grey50", linetype = 2))
obl_MVR_fit_bp2
```
This model identifies total number of federal agencies, the state/territory being Puerto Rico, and emergency declarations as significant predictors of obligated FEMA funding. It makes sense that an emergency declaration is a protective factor, as it is a smaller event than a major disaster declaration.

>>>>

Create box and whisker plot for the fit output.
```{r req SLR bp}
req_SLR_bp <- broom.mixed::tidy(req_SLR_fit) %>%
                dotwhisker::dwplot(dot_args = list(size = 2, color = "blue"),
                                   whisker_args = list(color = "blue"),
                                   vline = geom_vline(xintercept = 0, colour = "grey50", linetype = 2))
req_SLR_bp

#save image
req_SLR_bp_file = here("results","req_SLR_bp.png")
ggsave(filename = req_SLR_bp_file, plot = req_SLR_bp)
```

@@

#box and whisker plot for lm_fit2 significant predictors
req_MVR_fit_bp2 <- req_MVR_fit_sig %>%
                        dotwhisker::dwplot(dot_args = list(size = 2, color = "blue"),
                                           whisker_args = list(color = "blue"),
                                           vline = geom_vline(xintercept = 0, colour = "grey50", linetype = 2))
req_MVR_fit_bp2

#save file
req_MVR_bp_file = here("results","req_MVR_bp.png")
ggsave(filename = req_MVR_bp_file, plot = req_MVR_fit_bp2)
```
This model identifies total number of federal agencies, the state/territory being Puerto Rico, and emergency declarations as significant predictors of requested FEMA funding. It makes sense that an emergency declaration is a protective factor, as it is a smaller event than a major disaster declaration.

@@

Create box and whisker plot for the fit output.
```{r obl SLR bp}
obl_SLR_bp <- broom.mixed::tidy(obl_SLR_fit) %>%
                dotwhisker::dwplot(dot_args = list(size = 2, color = "blue"),
                                   whisker_args = list(color = "blue"),
                                   vline = geom_vline(xintercept = 0, colour = "grey50", linetype = 2))
obl_SLR_bp

#save image
obl_SLR_bp_file = here("results","obl_SLR_bp.png")
ggsave(filename = obl_SLR_bp_file, plot = obl_SLR_bp)
```

@@

#box and whisker plot for lm_fit2 significant predictors
obl_MVR_fit_bp2 <- obl_MVR_fit_sig %>%
                        dotwhisker::dwplot(dot_args = list(size = 2, color = "blue"),
                                           whisker_args = list(color = "blue"),
                                           vline = geom_vline(xintercept = 0, colour = "grey50", linetype = 2))
obl_MVR_fit_bp2

#save file
obl_MVR_bp_file = here("results","obl_MVR_bp.png")
ggsave(filename = obl_MVR_bp_file, plot = obl_MVR_fit_bp2)
```
This model identifies total number of federal agencies, the state/territory being Puerto Rico, and emergency declarations as significant predictors of obligated FEMA funding. It makes sense that an emergency declaration is a protective factor, as it is a smaller event than a major disaster declaration.

### 2021-11-13 code/processing_code/processing-populations.Rmd
Manuscript updates

## Introduction
This markdown imports the raw data and performs some cleaning as well as feature engineering. The raw data comes from  [the U.S> Department of Agriculture Economic Research Service](https://www.ers.usda.gov/data-products/county-level-data-sets/download-data/).

>>>>

## Introduction
This markdown imports the raw data and performs some cleaning as well as feature engineering. The raw data comes from  [the U.S. Department of Agriculture Economic Research Service](https://www.ers.usda.gov/data-products/county-level-data-sets/download-data/).

### 2021-11-13 products/manuscript/Manuscript_Disasters.Rmd
Manuscript updates

# Summary/Abstract
_Write a summary of your project._

@@

## General Background Information
_Provide enough background on your topic that others can understand the why and how of your analysis_

## Description of data and data source
_Describe what the data is, what it contains, where it is from, etc. Eventually this might be part of a methods section._

## Questions/Hypotheses to be addressed
_State the research questions you plan to answer with this analysis._

The funding process was further refined as part of the Sandy Recovery Improvement Act of 2013 as well as The Disaster Recovery Reform Act of 2019. These laws require FEMA to publicly provide the following information for each mission assignment:

When disasters occur, one of the most complicated and difficult challenges becomes financing the response and recovery efforts. In 1988, United States (US) President Ronald Reagan signed the Robert T. Stafford Disaster Relief and Emergency Act into law to formalize the financial responsibilities of each tier of US government. Under this law, when a disaster occurs, the affected area (local or state representatives) conducts a damage assessment to estimate the amount of support required from the federal government. That assessment goes through the state government to the Federal Emergency Management Agency (FEMA) to the Department of Homeland Security (DHS) to the US President, who ultimately decides whether the affected region should receive an emergency declaration, a major disaster declaration, or neither. FEMA then deploys resources from across the federal government to support the disaster region, the volume and type of which depend upon the declaration type. However, the federal government does not necessarily incur all of the cost of these resources. Under most circumstances, this pay structure more accurately reflects a reimbursement model; that is, the states or territories submit requests to the FEMA for reimbursement after the expenses have occurred. However, there is then a reconciliation process by which the total funds a state or territory receives is often different than the original request. The accumulation of this process, including characteristics of the incident itself, are extraordinarily complex, and therefore make it difficult to predict the cost and true impact.

Moreover, the combination of rapidly increasing inflation and worsening effects of climate change has caused disasters to become more frequent and devastating on an annual basis, and the financial impacts are becoming increasingly difficult to absorb within the United States (US) government infrastructure. It would greatly benefit the federal government to be able to predict the amount of funding required for a disaster based on parameters known at the onset of the incident, as it would allow for improved funds flow and budgeting practices. Moreover, it would benefit states and territories to know which parameters increase the amount of federal government support in the aftermath of a disaster. As such, this analysis will examine the use of machine learning models in predicting FEMA expenses for a federally-declared disaster. In particular, a decision tree, bagged tree, and random forest model will be used to predict obligated and requested FEMA funds.

@@

## Data import and cleaning
_Write code that reads in the file and cleans it so it's ready for analysis. Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts. If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g. as supplement in a paper). All materials, including files that contain code, should be commented well so everyone can follow along._
The data were obtained from the OpenFEMA Data Sets website as well as the US Department of Agriculture Economic Research Service website. The "Mission Assignment" data was cleaned to represent the requested and obligated FEMA funds for each disaster declaration for each state or territory. Feature engineering was performed to create summary variables for total number of federal agencies involved in resource assistance, the amount of funding requested by the state or territory from FEMA, the amount of funding allocated to the state or territory from FEMA, and the average cost share for FEMA of the resources provided.

The data obtained from the US Department of Agriculture Economic Research was utilized to generate state population estimates and the number of counties per state.

The data summarizing disaster declarations was cleaned to represent the unique declarations per state

## Exploratory analysis

_Use a combination of text/tables/figures to explore and describe your data. You should produce plots or tables or other summary quantities for the most interesting/important quantities in your data. Depending on the total number of variables in your dataset, explore all or some of the others. FIgures produced here might be histograms or density plots, correlation plots, etc. Tables might summarize your data._

_Continue by creating plots or tables of the outcome(s) of interest and the predictor/exposure/input variables you are most interested in. If your dataset is small, you can do that for all variables. Plots produced here can be scatterplots, boxplots, violinplots, etc. Tables can be simple 2x2 tables or larger ones._

_To get some further insight into your data, if reasonable you could compute simple statistics (e.g. t-tests, simple regression model with 1 predictor, etc.) to look for associations between your outcome(s) and each individual predictor variable. Though note that unless you pre-specified the outcome and main exposure, any "p<0.05 means statistical significance" interpretation is not valid._

A total 2,051 observations were included in the original dataset, but 20 were removed for having no concentration recorded for any of the radioisotopes in question.

Table \@ref(tab:isotope_table) describes the three radioisotopes captured in the data.

```{r isotopetable,  echo=FALSE}
#load data for entire document
#path to data
#note the use of the here() package and not absolute paths
data_location <- here::here("data","processed_data","processeddata.rds")
data_location2 <- here::here("data", "processed_data", "processeddatalong.rds")

#load data
mydata <- readRDS(data_location)
mydata_long <- readRDS(data_location2)

#define row labels
table1::label(mydata$Iodine131) <- "Iodine 131 (Bq/m^3)"
table1::label(mydata$Caesium134) <- "Caesium 134 (Bq/m^3)"
table1::label(mydata$Caesium137) <- "Caesium 137 (Bq/m^3)"

#now combine into table
table1::table1(~ Iodine131 + Caesium134 + Caesium137, data = mydata)
```

Figure \@ref(fig:radioisotopesdays_50) depicts the relationship between the first 50 days after the incident and the measured concentration of each of the three radioisotopes. Radioactive decay is often used as an example of exponential decay; however, this pattern is not represented with this data. This phenomenon can be explained by the fact that fallout radiation instead follows a Poisson distribution pattern, due to the geographical dispersion of the particles.

```{r radioisotopesdays_50,  fig.cap='Radioisotope concentrations in European countries by days after Chernobyl..', echo=FALSE}
knitr::include_graphics("../../results/radioisotopesdays_50.png")
```
Fallout radioisotope concentrations were collected from 96 different locations across 96 countries.

Figure \@ref(fig:i131samplelocations) depicts the relationship between time elapsed and the measured concentration of I-131 across the different countries. Figures \@ref(fig:cs134samplelocations) and \@ref(fig:cs137samplelocations) are identical representations but for Cs-134 and Cs-137 respectively.

```{r i131locations,  fig.cap='Concentration of I-131 in European countries by days after Chernobyl.', echo=FALSE}
knitr::include_graphics("../../results/i131samplelocations.png")
```

```{r cs134locations,  fig.cap='Concentration of Cs-134 in European countries by days after Chernobyl.', echo=FALSE}
knitr::include_graphics("../../results/cs134samplelocations.png")
```

```{r cs137locations,  fig.cap='Concentration of Cs-137 in European countries by days after Chernobyl.', echo=FALSE}
knitr::include_graphics("../../results/cs137samplelocations.png")
```

@@

# Discussion

@@

## Strengths and Limitations
_Discuss what you perceive as strengths and limitations of your analysis._

@@

_Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end_

This paper [@Leek2015a] discusses types of analyses.

Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header. Many more style files for almost any journal [are available](https://www.zotero.org/styles). You also specify the location of your bibtex reference file in the YAML. You can call your reference file anything you like, I just used the generic word `references.bib` but giving it a more descriptive name is probably better.

>>>>

# Summary/Abstract
Rapidly increasing inflation rates and worsening effects of climate change are dramatically increasing the financial impacts of disasters. Within the United States, the Federal Emergency Management Agency (FEMA) is responsible for providing federal resources to support states and territories during the response phase as well as tracking the financial costs. All tiers of government would benefit from knowing which disaster declaration characteristics would change the amount of financial support from FEMA. This analysis utilizes machine learning technology to predict the funds requested by states or territories from FEMA as well as the funds eventually obligated to states or territories from FEMA for a disaster declaration. In particular, three models are considered: decision tree, bagged tree, and random forest. Cross-validation and bootstrap resampling structures are utilized, depending on the model, and the fits of the models are evaluated using the root mean square error (RMSE) statistic. The bagged tree model employing bootstrap resampling best fit both the requested and obligated fund models; however, future studies should consider other machine learning models as well as further refinement of the tuning paramters for the ones considered in this analysis. The prediction capabilities created by such models have the potential to dramatically improve comprehension of the key factors affecting financial disaster relief support.

@@

On November 8, 1988, United States (US) President Ronald Reagan signed the Robert T. Stafford Disaster Relief and Emergency Act into law, which formalizes the disaster declaration process and specifies the response responsibilities for each tier of government. Under this law, when a disaster occurs, the affected area (local or state representatives) conducts a damage assessment to estimate the amount of support required from the federal government. That assessment goes through the state government to the Federal Emergency Management Agency (FEMA) to the Department of Homeland Security (DHS) to the US President, who ultimately decides whether the affected region should receive an emergency declaration, a major disaster declaration, or neither. FEMA then deploys resources from across the federal government to support the disaster region, the volume and type of which depend upon the declaration type. However, the federal government does not necessarily incur all of the cost of these resources. Under most circumstances, this pay structure more accurately reflects a reimbursement model; that is, the states or territories submit requests to the FEMA for reimbursement after the expenses have occurred. However, there is then a reconciliation process by which the total funds a state or territory receives is often different than the original request. The funding process was further refined as part of the Sandy Recovery Improvement Act of 2013 as well as The Disaster Recovery Reform Act of 2019. These laws require FEMA to publicly provide the information regarding federal government resources paid for by state and federal government funds for each disaster declaration. As a result, the OpenFEMA Data Sets website now exists and provides information summarizing disaster declarations and mission assignments. While this has improved the accessible information, little has been done to draw conclusions from these data for predicting the financial impacts of disasters.

It would greatly benefit the federal government to be able to predict the amount of funding required for a disaster based on parameters known at the onset of the incident, as it would allow for improved funds flow and budgeting practices. Moreover, it would benefit states and territories to know which parameters increase the amount of federal government support in the aftermath of a disaster. As such, this analysis will examine the use of machine learning models in predicting FEMA expenses for a federally-declared disaster. In particular, a decision tree, bagged tree, and random forest model will be used to predict requested and obligated FEMA funds.

@@

## Data import and cleaning
The data were obtained from the OpenFEMA Data Sets website as well as the US Department of Agriculture Economic Research Service website. The mission assignments data was cleaned to represent the requested and obligated FEMA funds for each disaster declaration for each state or territory. Feature engineering was performed to create summary variables for total number of federal agencies involved in resource assistance, the amount of funding requested by the state or territory from FEMA, the amount of funding allocated to the state or territory from FEMA, and the average cost share for FEMA of the resources provided. Similarly, the disaster declaration summaries were also summarized to represent the unique disaster declarations for each state, bounded to the same time frame as the mission assignments data. Feature engineering was performed to create variables for incident duration, response duration, month of declaration, and year of the incident. The population data from the US Department of Agriculture Economic Research Service was incorporated into the disaster summary data, after creating a variable for state. The mission assignment data and declaration data were then joined to create a comprehensive dataset that was used for analysis. The supplementary files contain the RStudio 4.1 code used to process the data. The following R packages were utilized for importing and cleaning the data: here, tidyverse, skimr, and lubridate.

## Exploratory analysis
The processed data were then explored to characterize and summarize the relevant features in RStudio 4.1 with the following packages: here, tidyverse, skimr, summarytools, ggplot2, car, gtsummary, scales, cowplot, and table1. The supplementary files contain the code that generates the following results.

Table \@ref(tab:table1) reflects the results of the exploratory data analysis. Between February 2012 and November 2021, there were 446 disaster declarations for states and territories. Funds requested by states or territories for FEMA ranged from -\$855,200.3 to \$4,388,540,486, with an average of \$29,673,687, and FEMA funds obligated to the states and territories after reconciliation ranged from -\$855,200.3 to \$440,842,4297 with an average of \$29,744,109.

```{r summarytable,  echo=FALSE, message = FALSE}
readRDS("../../results/table1.rds")
```

Figure \@ref(fig:incidentyears) shows the number of disaster declarations per year, and Figure \@ref(fig:funds-years) shows the total FEMA funds (both requested and obligated) by states per year. The substantial peak in 2020 corresponds to the COVID-19 pandemic, and the one in 2017 represents the hurricane season with Hurricanes Irma, Harvey, and Maria. These characteristics can also be seen in Table \@ref(tab:decs-year-3), which represents FEMA Funding for the 3 Costliest Years (2012 - 2021).

```{r incidentyears,  fig.cap='Number of disaster declarations per year 2012 - 2021', echo=FALSE, message = FALSE}
knitr::include_graphics("../../results/incidentyear.png")
```

```{r funds-years,  fig.cap='State or territory requested FEMA funds and obligated FEMA funds by year 2012 - 2021', echo=FALSE, message = FALSE}
knitr::include_graphics("../../results/funds-years.png")
```

```{r decs-year-3,  echo=FALSE, message = FALSE}
readRDS("../../results/decs-year-3.Rds")
```

Figure \@ref(fig:incidentmonth) shows the number of disaster declarations per month, and Figure \@ref(fig:funds-month) shows the total FEMA funds (both requested and obligated) by states per month. The substantial peak in January likely corresponds to the COVID-19 pandemic, and the ones in August, September, and October represent the peak hurricane season. These characteristics can also be seen in Table \@ref(tab:decs-month-3), which represents FEMA Funding for the 3 Costliest Months.

```{r incidentmonth,  fig.cap='Number of disaster declarations per month for years 2012 - 2021', echo=FALSE}
knitr::include_graphics("../../results/incidentmonth.png")
```

```{r funds-month,  fig.cap='State or territory requested FEMA funds and obligated FEMA funds by month for 2012 - 2021', echo=FALSE}
knitr::include_graphics("../../results/funds-month.png")
```

```{r decs-month-3,  echo=FALSE, message = FALSE}
readRDS("../../results/decs-month-3.Rds")
```

All 50 states as well as Washington, DC were included in the final, combined dataset. Table \@ref(tab:decs-state-5) provides the five most disaster-prone states and summarizes their requested FEMA funding and received FEMA funding over the study period.
```{r decs-state-5,  echo=FALSE, message = FALSE}
readRDS("../../results/decs-state-5.Rds")
```

Incident duration ranged from zero days to 660 days (or almost two years), as seen by Figure \@ref(fig:funds-dur). The incident duration of 660 days likely corresponds to the COVID-19 pandemic, which is an ongoing incident of almost two years. Further analysis of all variables can be found in the supplementary information.

```{r funds-dur,  fig.cap='Incident duration for requested FEMA funds and obligated FEMA funds under $1 billion for 2012 - 2021', echo=FALSE}
knitr::include_graphics("../../results/funds-dur.png")
```

Prior to fitting the data to machine learning models, standard simple and multivariable regression models were fitted using the tidymodels framework in RStudio 4.1. Both outcomes of interest were fitted, using incident duration as the primary predictor of interest. Figure \@ref(fig:req_MVR_bp) and \@ref(fig:obl_MVR_bp) show the significant variables fitted in the multivariable regression models for requested and obligated FEMA funds, respectively.

```{r req_MVR_bp,  fig.cap='Significant predictors in multivariate regression model for requested FEMA funds', echo=FALSE}
knitr::include_graphics("../../results/req_MVR_bp.png")
```

```{r obl_MVR_bp,  fig.cap='Significant predictors in multivariate regression model for obligated FEMA funds', echo=FALSE}
knitr::include_graphics("../../results/obl_MVR_bp.png")
```

Table \@ref(tab:lm-eval) shows the comparisons of the models, using the root mean square error (RMSE) and r-squared statistics. The requested and obligated funding models performed almost identically. None of the models fit the data well, which suggests more sophisticated modeling practices are necessary to adequately characterize the data.

```{r lm-eval,  echo=FALSE, message = FALSE}
resulttable2=readRDS("../../results/lm-eval.rds")
knitr::kable(resulttable2, caption = 'Linear model comparisons')
```

@@

Three models were considered for the prediction analysis: decision tree, bagged tree, and random forest, and each were fit for both outcomes of interest (e.g. requested and obligated funds). Prior to fitting the models, the data were subset based on outcome of interest, and some feature engineering was performed. Due to the limited values in certain categories, the state variable was summarized into the ten FEMA regions and the least frequent disaster types were added to the "Other" category. The data were split such that 66.7% were used for training the models and the remaining 33.3% were used for testing the final chosen model, stratified by the incident type. Cross-validation was used for the decision tree and random forest models, using a five-fold, five-repeat resampling structure, also stratified by the incident type. The bagged tree model instead implemented a bootstrap resampling structure, repeated ten times, again stratified by the incident type. The models were then compared based on the root mean square error (RMSE), the standard metric for regression models. The analysis was completed in RStudio 4.1 with the following packages: here, tidyverse, tidymodels, skmir, broom.mixed, rpart.plot, vip, glmnet, doParallel, ranger, and baguette.

The identified decision tree model for each outcome is illustrated by Figures  \@ref(fig:DT-plot) and  \@ref(fig:DT-plot-obl). Both outcomes identified the same decision point features: total number of federal agencies involved in the response, duration of the incident, state population, state counties, and whether the individual / housing assistance program was awarded.

```{r DT-plot,  fig.cap='Selected decision tree for predicting requested FEMA funds', echo=FALSE}
knitr::include_graphics("../../results/DT-plot.png")
```

```{r DT-plot-obl,  fig.cap='Selected decision tree for predicting obligated FEMA funds', echo=FALSE}
knitr::include_graphics("../../results/DT-plot-obl")
```

The most important variables in the identified random forest model for each outcome is illustrated by Figures  \@ref(fig:RF-plot) and  \@ref(fig:RF-plot-obl). Both outcomes identified the same important features: incident duration, total number of agencies involved, individual/housing assistance program awardance, and incident month. Interestingly, however, the obligated funds model includes days of the entire response as the fifth most important feature, whereas the requested funds model includes it as the eleventh most important feature. Unfortunately, the tidymodels framework does not support visualization of the bagged tree model.

```{r RF-plot,  fig.cap='20 most important variables for the random forest model for requested FEMA funds', echo=FALSE}
knitr::include_graphics("../../results/RF-plot.png")
```

```{r RF-plot-obl,  fig.cap='20 most important variables for the random forest model for obligated FEMA funds', echo=FALSE}
knitr::include_graphics("../../results/RF-plot-obl")
```

Table \@ref(tab:ML-model-comp) shows the model comparisons for estimating requested FEMA funding, and Table \@ref(tab:ML-model-comp-obl) similarly shows the model comparisons for estimating obligated FEMA funding. In both outcomes, the bagged tree had the lowest RMSE; however, the standard error estimate of the RMSE was higher in the bagged tree models than for the other models.

```{r ML-model-comp,  echo=FALSE, message = FALSE}
resulttable=readRDS("../../results/ML-model-comp.rds")
knitr::kable(resulttable, caption = 'ML model comparison for requested FEMA funding')
```

```{r ML-model-comp-obl,  echo=FALSE}
resulttable1=readRDS("../../results/ML-model-comp-obl.rds")
knitr::kable(resulttable1, caption = 'ML model comparison for obligated FEMA funding')
```

The bagged tree model was still chosen as the best model of the three, and it was fitted to the testing data sample. The RMSEs for the testing and training datasets were substantially different. Figure \@ref(fig:BT-pred-act-test) shows the predicted and actual requested funds values, and Figure \@ref(fig:BT-pred-act-test-obl) does the same for obligated FEMA funds. The supplementary information contains the code for the models as well as all diagnostic plots.
```{r BT_pred_act_test,  fig.cap='Predicted vs actual requested FEMA funds for the bagged tree fit', echo=FALSE}
knitr::include_graphics("../../results/BT-pred-act-test.png")
```

```{r BT_pred_act_test-obl,  fig.cap='Predicted vs actual obligated FEMA funds for the bagged tree fit', echo=FALSE}
knitr::include_graphics("../../results/BT-pred-act-test-obl.png")
```

# Discussion
This analysis applies machine learning modeling techniques to predict the requested and obligated funds from FEMA in the aftermath of a disaster.

@@

## Strengths and Limitations
This analysis shows promise but requires further tuning and validation. THe usefulness of the model is limited by the broad variance present in the relatively small dataset, which potentially leads to biases. Moreover, the COVID-19 pandemic created extensive outliers in the dataset, which further complicate the prediction algorithms. While it is tempting to remove the outlier from a statistics perspective, it makes less sense scientifically, as the outlier disasters are often the ones where funding prediction would be the most useful.

@@
